Linting Linearity in Core/System FC

### 2022-11-24 proposal.tex
Apply text patches

Since Linear Haskell's\cite{} publication and implementation release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types into a mainstream pure and lazy functional language. In concrete,
function arrows can be annotated with a multiplicity. If a function is linear it
must consume its argument exactly once if it is consumed exactly once, for some
definition of \emph{consume} that I should revise here.

Linearly typed programs are proven/checked to be linear by the type system, so
the addition of linear types evidently implied changing the type checker to
support linearity. There exist, however, two distinct type checkers in Haskell.
The first is run on the surface language, i.e. the Haskell we write, and is a
big and complex type checker that now also supports typing linearity. The second
is run on the intermediate language \emph{Core} that we obtain from desugaring
Haskell.

\emph{Core} is a much smaller language than the whole of Haskell (even though we
can compile the whole of Haskell to it!) and the type checker for it is small
and fast due to \emph{Core} being explicitly typed and having a small abstract syntax
tree. This type checker (called \emph{Lint}) gives us guarantees of correctness
(i.e. sanity) in face of all the complex transformations a Haskell program
undergoes such as desugaring and optimising transformations because the linter
is always run on the resulting code before being compiled to untyped machine
code.

@@

We want Core to give us guarantees about our desugaring and transformations with
regard to linearity too -- a linearly typed Core will ensure that linearly-typed
programs remain correct (i.e.  linearity is preserved) after desugaring and all
GHC transformations (optimisations shouldn't destroy linearity).

Core is already annotated with linearity, but \textbf{we currently ignore it}!
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far from
trivial, and the implemented formal system cannot accomodate various
optimising transformations and from them resulting programs, that seemingly
violate linearity, but which do preserve it semantically. Therefore, we reject
various valid programs with regard to linearity after desugaring: the current
solution, because disabling optimisations can inccur great performance costs, is
to disable the linear linter. We believe that GHC's transformations are correct,
and it is the linear type system that can't accommodate the resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and the
% linter rejects valid programs which is quite undesirable.

@@

We propose to formalise a type system that is able to type check linear
\emph{Core} at all points in the core pipeline and then to implement it into
GHC. Proposal... \todo{proposal paragraph}

@@

The first set of problems appears in the core-to-core optimisation passes. GHC
applies many optimising transformations to \emph{Core} and we believe those
transformations preserve linearity. However, our linear type system cannot check
that they indeed preserve linearity.

The simpler examples come from straightforward and common optimising
transformations. Then we have recursive let definitions that don't accommodate
linearity even though it might converge to only use the value once. Finally, we
have the empty case expression introduced with the \emph{EmptyCase} language
extension that we currently can't typecheck either.

@@

The solution with the unifying case binder usage idea means annotating the case
binder with its usage environment. For $True$ and $False$ (and other data
constructors without arguments) the usage environment is always empty (using
them doesn't entail using any other variable), meaning we can always use the
case binder instead of the actual constant constructor without issues.
%
In concrete, if we had $w1$ instead of the second occurrence of $x$, we'd have
an empty usage environment for $w1$ in the $False$ branch ($w1 :_\emptyset
Bool$) and upon using $w1$ we wouldn't use any extra resources

>>>>

Since Linear Haskell's\cite{} publication and implementation release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types into a mainstream pure and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once (its arguments have multiplicity of one), for some
definition of \emph{consume} that I should revise here.

Linearly typed programs are proven/checked to be linear by the type system, so
the addition of linear types evidently implied changing the type checker to
support linearity. There exist, however, two distinct type checkers in Haskell.
The first is run on the surface language, i.e. the Haskell we write, and is a
very complex type checker that now also supports typing linearity. The second is
run on the intermediate language \emph{Core} that we obtain from desugaring
Haskell.

\emph{Core} is a much smaller language than the whole of Haskell (even though we
can compile the whole of Haskell to it!) and the type checker for it is small
and fast due to \emph{Core} being explicitly typed and much closer to a
polymorphic lambda calculus than its surface-level counterpart. This type
checker (called \emph{Lint}) gives us guarantees of correctness in face of all
the complex transformations a Haskell program undergoes, such as desugaring and
core-to-core passes, because the linter is always run on the resulting code
before being compiled to untyped machine code.

@@

We want Core to give us guarantees about our desugaring and transformations with
regard to linearity too -- a linearly typed Core will ensure that linearly-typed
programs remain correct (linearity is preserved) after desugaring and all GHC
transformations (optimisations shouldn't destroy linearity).

Core is already annotated with linearity, but the \textbf{linter currently
ignores it}! In spite of the strong formal foundations of linear types driving
the implementation, their interaction with the whole of GHC is still far from
trivial, and the implemented type system cannot accomodate various optimising
transformations and, from them resulting, programs that seemingly violate
linearity but which do preserve it semantically; for example, laziness makes it
harder to reason about linearity. Therefore, we reject various valid programs
with regard to linearity after desugaring: the current solution, because
disabling optimisations can inccur great performance costs, is to disable the
linear linter. However, we believe that GHC's transformations are correct, and
it is the linear type system that can't accommodate the resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and
% the
% linter rejects valid programs which is quite undesirable.

@@

We propose to extend Core's formal type system with more linearity information
to be able to accommodate linear programs as seen in Core after optimising
transformations, prove its soundness/type safety, validate core-to-core
optimisation passes, and ultimately implement it into GHC's linter. Concretely,
we'll extend the linear Core system with rules based on usage annotations for
let, letrec and case binder bound variables, possibly a new kind of coercion for
multiplicities, and eventually other rules.
%
The ultimate measure of success is the -dlinear-core-lint flag that, right now,
when enabled, rejects many linearly valid programs in the linter. Ideally, by
the end of our research and implementation, this flag could be enabled by
default and accommodate all existing transformations. Realistically, we want to
formally (in the type system) accept as many diverse transformations as possible
while still preserving linearity, even if we are unable to fix them all.

@@

The first set of problems appears in the core-to-core optimisation passes. GHC
applies many optimising transformations to \emph{Core} and we believe those
transformations preserve linearity. However, Core's linear type system cannot check
that they indeed preserve linearity.

The simpler examples come from straightforward and common optimising
transformations. Then we have recursive let definitions that don't accommodate
linearity even though it might converge to only use the value once. Finally, we
have the empty case expression introduced with the \emph{EmptyCase} language
extension that Core currently can't typecheck either.

@@

The solution with the unifying case binder usage idea means annotating the case
binder with its usage environment. For $True$ and $False$ (and other data
constructors without arguments) the usage environment is always empty (using
them doesn't entail using any other variable), meaning we can always use the
case binder instead of the actual constant constructor without issues.
%
Concretely, if we had $w1$ instead of the second occurrence of $x$, we'd have
an empty usage environment for $w1$ in the $False$ branch ($w1 :_\emptyset
Bool$) and upon using $w1$ we wouldn't use any extra resources
